The host galaxy of the short GRB~111117A at $z = 2.211$\thanks{Based on observations collected at the European Southern Observatory, Paranal, Chile, Program ID: 088.A-0051 and 091.D-0904.}

### 2017-06-11 GRB111117A.tex
Addressing Steves comments
Changing comparison sample to D’Avanzo for completeness. Closes #15

The host lies in the brighter end of the expected host brightness distribution
at $z = 2.211$, and is actively forming stars. Using the host as a benchmark for
redshift determination, we find that less than 55 per cent of all sGRB redshifts
should be missed due to host faintness at $z\sim2$. The high redshift of
GRB~111117A is evidence against a lognormal delay-time model for sGRBs through
the predicted redshift distribution of sGRBs, which is very sensitive to
high-$z$ sGRBs.

@@

There is mounting evidence that short-duration $\gamma$-ray bursts
come from the merger of NSs, either with another NS, or a black
hole, due to their apparent association with kilonovae \citep{Barnes2013a,
	Tanvir2013b, Yang2015, Jin2016, Rosswog2016}. The absence of associated
supernovae in deep searches \citep[e.g.][]{Hjorth2005a,Fox2005,Hjorth2005b}
supports this idea and distinguishes the physical origin of sGRBs from their
long-duration counterparts. %\citep[albeit see also][]{Fynbo2006b, Valle2006, Gal-Yam2006}.

@@

Spectroscopic observations were carried out using the cross-dispersed echelle
spectrograph, VLT/X-shooter \citep{Vernet2011}, at four seperate epochs. The
burst was observed 38 hours after the Burst Alert Telescope (BAT) trigger under
ESO programme 088.A-0051 (PI: Fynbo) and again later under ESO programme
091.D-0904 (PI: Hjorth). X-shooter covers the wavelength range from 3000~\AA~to
24\,800~\AA~(21\,000~\AA~when the $K$-band blocking filter is used) across three
spectroscopic arms. We carried out the bias-correction, flat-fielding, order
tracing, wavelength calibration, rectification, and flux calibration using the
VLT/X-shooter pipeline version 2.8.4 \citep{Modigliani2010}.
%Because the spectra are curved across each detector, a
% rectification algorithm
%is employed which introduces correlations between neighboring pixels. We select
%a pixel-scale of 0.2/0.2/0.6 \AA/pix for the UVB/VIS/NIR arm to minimize the
%degree of correlation while conserving the maximal resolution.
The observations are combined and extracted using scripts described in Selsing
et al. 2017 (in prep.) and available
online\footnote{\url{https://github.com/jselsing/XSGRB_reduction_scripts}}. 
%The
%signal-to-noise ratio of the continuum in the near-infrared arm is too low to use the
%optimal extraction algorithm \citep{Horne1986} and therefore the extraction is
%carried out with a simple aperture.
An overview of the spectroscopic observations is given in
Table~\ref{tab:spec_overview}, and the slit position is shown in
Fig.~\ref{fig:spec_setup}. We show the extracted spectrum in Fig.~\ref{fig:SED}.

@@

A single sGRB at high redshift does little in terms of constraining the redshift
distribution of sGRBs. In particular, other sGRB hosts could be missed because
they are intrinsically fainter and thus this high-$z$ event is only detected due
to the brightness of it's host. \citet{Berger2014} compiled a sample of sGRB
host luminosities, normalized by the characteristic galaxy luminosity at their
respective redshift, $L_B/L^{\star}_{B}$. Out of the 39 hosts in the sample, 26
(66 per cent) have measured redshifts. To convert the SED-inferred $M_B$ of GRB~111117A
to $L_B/L^{\star}_{B}$, we use the characteristic absolute $B$-band magnitude of
the Schechter function for blue galaxies ($U - V < 0.25$) in the redshift window
$2.0 \leq z \leq 2.5$ from \citet{Marchesini2007}. For GRB~111117A, we obtain
$L_B/L^{\star}_{B} = 1.2$, which is brighter than 70 per cent of the hosts in
\citet{Berger2014} with measured $L_B/L^{\star}_{B}$.

If we assume that we are able to obtain emission-line redshifts from hosts with
$R < 25$~mag \citep{Kruhler2012}, then we would have missed around 30 per cent
(8 out of 26 from the sample of \citealt{Berger2014} with measured
$L_B/L^{\star}_{B}$), if they were at the redshift of GRB~111117A. Because the
average SFR of galaxies hosting lGRBs is higher than for galaxies hosting sGRBs,
the fraction of missed burst redshifts is likely higher although the cosmic SFR
evolution could play a role in improving redshift determinability.

A fraction of the bursts missing redshift are host-less and is therefore likely
at moderate redshifts \citep{Tunnicliffe2014}, but should some of the remainder
be at high redshift, the missed fraction will increase. If we assume that
\textit{all} the bursts that are missing redshifts in \citet{Berger2014} are at
high-$z$ and missed due to host faintness, 22 out of 39 hosts (55 per cent)
would be missed at $z = 2.211$. This serves as an upper limit on the fraction of
missed burst at high-$z$ and illustrates that we are likely \textit{not} missing a
large fraction of sGRBs redshift at $z \approx 2$ due to host faintness.

@@

At $z = 2.211$, the age of the universe is 3 Gyr. If the progenitor systems of
sGRBs are the merger of two NSs, this sets a hard upper limit to the coalescence
timescale for such a system. In the absence of other mechanisms, the timescale
of the orbital decay of the system is set by the energy loss due to
gravitational waves, which in turn is set by the mass of the constituent compact
objects and the separation of the two \citep{Postnov2014}. If we assume that the
formation timescale of the first galaxies is short compared to the time since
the Big Bang \citep{Richard2011}, and if we assume a mass of 1.4~$M_\odot$ for
each of the NSs at the time of system formation, this places a hard upper limit
on the initial separation, of $a_0 < 3.2~R_\odot$.

In practice most NS-NS binaries will be eccentric at formation because of the SN
natal kicks, and this can be quite a big effect on merger times, although
eccentric systems normally circularise first. It could also be a BH-NS system,
and as the merger time scales roughly as $M^3$ the mass here almost as big an
effect as the separation ($a^4$).

@@

Although a single burst carries little leverage in terms of constraining the
redshift distribution of sGRB, the high redshift of GRB~111117A needs to be
accommodated in progenitor models. A lognormal delay time model predicts a very low volumetric
density of bursts at $z = 2.211$, whereas a power law delay time model peaks
near GRB~111117A. If more sGRBs are at similarly high redshifts, but are missed due to
the faintness of their hosts, a lognormal delay time model will be disfavored.
Compared to a sample of sGRB hosts, GRB~111117A is more luminous than 70 per
cent of the sample with measured luminosities. Conservatively, for 55 per
cent of the sGRB hosts, we would be unable to determine a redshift should they be at 
a similar redshift as GRB~111117A. This implies that we are \textit{not} missing a large
fraction of the sGRBs at $z \sim 2$.

>>>>

The host lies in the brighter end of the expected host brightness distribution
at $z = 2.211$, and is actively forming stars. Using the host as a benchmark for
redshift determination, we find that less than 64 per cent of all sGRB redshifts
should be missed due to host faintness at $z\sim2$. The high redshift of
GRB~111117A is evidence against a lognormal delay-time model for sGRBs through
the predicted redshift distribution of sGRBs, which is very sensitive to
high-$z$ sGRBs.

@@

There is mounting evidence that short-duration $\gamma$-ray bursts
come from the merger of NSs, either with another NS, or a black
hole, due to their apparent association with kilonovae \citep{Barnes2013a,
	Tanvir2013b, Yang2015, Jin2016, Rosswog2016}. The absence of associated
supernovae in deep searches \citep[e.g.][]{Hjorth2005a,Fox2005,Hjorth2005b, Kann2011}
supports this idea and distinguishes the physical origin of sGRBs from their
long-duration counterparts. %\citep[albeit see also][]{Fynbo2006b, Valle2006, Gal-Yam2006}.

@@

Spectroscopic observations were carried out using the cross-dispersed echelle
spectrograph, VLT/X-shooter \citep{Vernet2011}, at four seperate epochs. The
burst was observed 38 hours after the Burst Alert Telescope (BAT) trigger under
ESO programme 088.A-0051 (PI: Fynbo) and again later under ESO programme
091.D-0904 (PI: Hjorth). X-shooter covers the wavelength range from 3000~\AA~to
24\,800~\AA~(21\,000~\AA~when the $K$-band blocking filter is used) across three
spectroscopic arms. We carried out the bias-correction, flat-fielding, order
tracing, wavelength calibration, rectification, and flux calibration using the
VLT/X-shooter pipeline version 2.8.4 \citep{Modigliani2010}.
%Because the spectra are curved across each detector, a rectification algorithm
%is employed which introduces correlations between neighboring pixels. We select
%a pixel-scale of 0.2/0.2/0.6 \AA/pix for the UVB/VIS/NIR arm to minimize the
%degree of correlation while conserving the maximal resolution.
The observations are combined and extracted using scripts described in Selsing
et al. 2017 (in prep.) and available
online\footnote{\url{https://github.com/jselsing/XSGRB_reduction_scripts}}. 
%The signal-to-noise ratio of the continuum in the near-infrared arm is too low
%to use the optimal extraction algorithm \citep{Horne1986} and therefore the
%extraction is carried out with a simple aperture.
An overview of the spectroscopic observations is given in
Table~\ref{tab:spec_overview}, and the slit position is shown in
Fig.~\ref{fig:spec_setup}. We show the extracted spectrum in Fig.~\ref{fig:SED}.

@@

A single sGRB at high redshift does little in terms of constraining the redshift
distribution of sGRBs. In particular, other sGRB hosts could be missed because
they are intrinsically fainter and thus this high-$z$ event is only detected due
to the brightness of it's host. \citet{Berger2014} compiled a sample of sGRB
host luminosities, normalized by the characteristic galaxy luminosity at their
respective redshift, $L_B/L^{\star}_{B}$. To convert the SED-inferred $M_B$ of
GRB~111117A to $L_B/L^{\star}_{B}$, we use the characteristic absolute $B$-band
magnitude of the Schechter function for blue galaxies ($U - V < 0.25$) in the
redshift window $2.0 \leq z \leq 2.5$ from \citet{Marchesini2007} and obtain
$L_B/L^{\star}_{B} = 1.2$.

Using the complete, flux-limited selection of bursts from \citet{DAvanzo2014a},
excluding GRB~111117A and the non-sGRB GRB~090426, we have a statistically
homogeneous sample from which we can address the implications of the redshift of
GRB~111117A. Out of the 14 hosts in the sample, 10 (71 per cent) have measured
redshifts and $L_B/L^{\star}_{B}$. Of the complete sample, the host of
GRB~111117A  is brighter than 80 per cent of the hosts with measured
$L_B/L^{\star}_{B}$. If we conservatively assume that \textit{all} the host
missing $L_B/L^{\star}_{B}$ are brighter than the host of GRB~111117A, the host
still brighter than > 60 per cent of sGRB hosts. For all 26 hosts with
$L_B/L^{\star}_{B}$ from \citet{Berger2014}, the host of GRB~111117A is brighter
than 73 per cent.

If we assume that we are able to obtain emission-line redshifts from hosts with
$R < 25$~mag \citep{Kruhler2012}, then we would have missed around 50 per cent
of the redshifts (5 out of 10 hosts), if they were at the redshift of
GRB~111117A due to the hosts being too faint. The corresponding number is around
30 per cent (8 out of 26) from the full sample of \citet{Berger2014}, reflecting
the lower mean $L_B/L^{\star}_{B}$ of the complete sample. Because the average
SFR of galaxies hosting lGRBs is higher than for galaxies hosting sGRBs, the
fraction of missed burst redshifts is likely higher although the cosmic SFR
evolution could play a role in improving redshift determinability.

A fraction of the bursts missing redshift are host-less and is therefore likely
at moderate redshifts \citep{Tunnicliffe2014}, but should some of the remainder
be at high redshift, the missed fraction will increase. If we assume that
\textit{all} the bursts that are missing redshifts are at high-$z$ and missed
due to host faintness, 9 out of 14 hosts in the complete sample (64 per cent)
would be missed at $z = 2.211$. This serves as an upper limit on the fraction of
missed burst redshifts at high-$z$. The corresponding fraction of missed bursts
is 0.55 (22 out of 39 hosts) using the full sample from \citet{Berger2014}.

@@

At $z = 2.211$, the age of the universe is 3 Gyr. If the progenitor systems of
sGRBs are the merger of two NSs, this sets a hard upper limit to the coalescence
timescale for such a system. In the absence of other mechanisms, the timescale
of the orbital decay of the system is set by the energy loss due to
gravitational waves, which in turn is set by the mass of the constituent compact
objects, the eccentricity of the orbit and the separation of the two \citep{Postnov2014}. If we assume that the
formation timescale of the first galaxies is short compared to the time since
the Big Bang \citep{Richard2011}, and if we assume a mass of 1.4~$M_\odot$ for
each of the NSs in a circular orbit at the time of system formation, this places
a hard upper limit on the initial separation, of $a_0 < 3.2~R_\odot$.

In practice most NS-NS binaries will be eccentric at formation because of the SN
natal kicks, although eccentric systems normally circularise first. For more
eccentric orbits, the coalescence timescale decreases, leading to a decrease in
the constraint on the initial separation and larger initial separations. As
noted by \citet{Postnov2014}, it takes eccentricities > 0.6 to significantly
shorten the merger time. 
%It could also be a BH-NS system, and as the merger time scales roughly as $M^3$
%the mass here almost as big an effect as the separation ($a^4$).

@@

Although a single burst carries little leverage in terms of constraining the
redshift distribution of sGRB, the high redshift of GRB~111117A needs to be
accommodated in progenitor models. A lognormal delay time model predicts a very
low volumetric density of bursts at $z = 2.211$, whereas a power law delay time
model peaks near GRB~111117A. If more sGRBs are at similarly high redshifts, but
are missed due to the faintness of their hosts, a lognormal delay time model
will be disfavored. Compared to a sample of sGRB hosts, GRB~111117A is more
luminous than 80 per cent of the sample with measured luminosities.
Conservatively, for 64 per cent of sGRB hosts, we would be unable to determine a
redshift should they be at a similar redshift as GRB~111117A.
